Efficient static analysis of Marlowe contracts

### 2020-05-21 papers/static_analysis2020/paper.tex
Address some reviews by @simonjohnthompson

\maketitle
\begin{abstract}
SMT solvers can verify properties very efficiently, and they offer increasing flexibility on the ways those properties can be described. However, the way in which properties are described can affect the computational cost of verifying them considerably. In addition, it is quite hard to predict how different changes will affect the computational cost.

@@

Thanks to static analysis, we can automatically check beforehand whether
payments promised by a Marlowe contract can be fulfilled in every
possible execution of the contract. For example, if a Marlowe contract
has passed the static analysis, we will have a very high assurance
that whenever the contract says we will be paid a certain amount of
money, the contract will indeed have enough money to make such payment.

@@

This paper contributes a number of approaches that can be used when
optimising static analysis, examples extracted from a case-study where
we applied these approaches, and an overview of the techniques used
to test the correctness of the optimisations.

The case study presented is that of the optimisation of Marlowe's
static analysis. During this work, the techniques applied allowed
us to reduce the analysis time of an implementation that followed
the semantics closely and took a couple of minutes to analyse contracts
of a few kilobytes, to one where the same contract would take less
than a second, and where a four-person crowdfunding contract that
when fully expanded occupies about 19 megabytes, can be analysed in
around 10 minutes using Z3. In the original implementation, we were
not able to analyse the 19 megabyte contract because it ran out of
memory after half an hour.

@@

modifications are local and can be done without fundamentally changing
the implementation. We consider three main ideas:
\begin{itemize}
\item \textbf{Removing unnecessary parts from the analysis}. If they do not affect
the property that is being verified then we can just remove them from
the implementation and reduce its complexity.
\item \textbf{Avoiding high level abstractions}. High level abstractions aid reasoning
and avoid errors, but also introduce complexity that is not necessarily
needed.
\item \textbf{Reducing search space by normalising parameters}. If there are several
ways of representing some inputs, and the different representations
have no impact on the analysis, we can remove all but one when
implementing static analysis.
\end{itemize}

@@

modifications are more fundamental approaches that require considerable
changes to the structure of the implementation. We consider two main ideas:
\begin{itemize}
\item \textbf{Reducing search space by using normalised execution paths relevant
to the property}. Instead of using search space to model inputs, we
use it to model possible executions, and we ensure that we only
represent each possible group of equivalent executions once, i.e:
executions with the same normal form.
\item \textbf{Minimizing the representation of inputs and outputs}. We can benefit
from encoding inputs and outputs as concisely as possible, and discarding
information that we can infer in other ways.
\end{itemize}
In the following sections, we introduce the semantics of Marlowe as
a case study (Section~\ref{sec:Marlowe-intro}), we go through a
general approach to static analysis (Section~\ref{sec:Making-Marlowe-symbolic}),
then we explore both lightweight (Section~\ref{subsec:Lightweight-modifications})
and heavyweight (Section~\ref{subsec:Heavyweight-modifications})
optimisation techniques in more detail, and we illustrate them with
examples of how they apply to Marlowe static analysis. In Section~\ref{sec:Testing},
we briefly discuss some ways in which property based testing can be
used to address potential problems that may arise from the implementation
of the heavyweight optimisations described in the previous section.
In Section~\ref{sec:Measurements}, we present empirical results that
show the effects of heavyweight optimisations on the execution time
and memory usage of Marlowe's static analysis.

@@

Marlowe contracts are able to receive payments, store money and tokens,
ask for input from the participants, and redistribute stored money
and tokens back among participants. The decision about which and when of
these actions are carried out is taken by the contract on the basis
of the information that is available at each point in time. Participants
may correspond to either individual public keys or to tokens (\texttt{Role}s).
In turn, \texttt{Role}s may be controlled by separate independent contracts.

The Marlowe language is written as a set of mutually recursive Haskell
data types. Marlowe contracts regulate interactions between a finite number
of participants, determined before the contract is deployed.

@@

There are 5 constructs of type \texttt{Contract}:
\begin{itemize}
\item \texttt{Close} -- signals the end of the life of a contract. Once
it is reached, all the money and tokens stored in the accounts of
the contract will be refunded to the owner of each of the respective
accounts.
\item \texttt{If} -- immediately decides how the contract must continue
from between two possibilities and depending on a given Boolean condition
that we call an \texttt{Observation}. The value of an \texttt{Observation}
can depend on previous choices made by participants, on amounts of money
and tokens remaining, or other factors.
\item \texttt{Let} -- immediately stores a \texttt{Value} for later use.
Expressions of type \texttt{Value} in Marlowe are evaluated to integer
values and they depend on information available to the contract at the
time of evaluation. For example, a \texttt{Let} construct can be used to
store the minimum slot of the slot interval of the current transaction,
or the amount of money available in a particular account at a given point
in time, for use at a later stage of the contract execution.
\item \texttt{When} -- waits for an external input. The \texttt{When} construct
also specifies a timeout slot: after this slot has been reached, the
\texttt{When} construct expires, and no longer accepts any input.
There are three types of input:
\begin{itemize}
\item \texttt{Deposit} -- Waits for a participant to deposit an amount
of money or tokens (specified as a \texttt{Value}) in the contract.
The amount is added to one of the accounts of the contract.
\item \texttt{Choice} -- Waits for a participant to make a choice. A choice
is represented as an integer number from a set specified by the contract.
\item \texttt{Notify} -- Waits for an \texttt{Observation} to be true.
Because contracts are reactive (they cannot initiate transactions),
it is necessary for an external actor to \texttt{Notify} the contract
that the \texttt{Observation} has become true. However, this notification
does not have to be carried out by a participant, anybody can do it.
\end{itemize}
\item \texttt{Pay} -- immediately makes a payment between accounts of the
contract, or from an account of the contract to a given participant.
The amount transferred is specified as a \texttt{Value}.
\end{itemize}

@@

As we mentioned in Section~\ref{subsec:Structure-of-Marlowe}, Marlowe
contracts are passive: their code is executed as part of the validation of
transactions that are submitted to the blockchain. These transactions
need to be submitted by participants or their representatives (e.g.
user wallets) and their validation is carried out atomically and
deterministically.

Each transaction may include a list of inputs, a set of signatures,
minimum and maximum slot numbers, a set of input UTxOs (incoming money
and tokens), and a set of outputs (outgoing money and tokens). For
a transaction to be valid in Marlowe, the transaction must have the
same effect for every slot within that slot range (be deterministic).
For example, if a transaction has a minimum slot number which is before
a timeout, and a maximum slot which after a timeout then the transaction
will fail with the error \texttt{AmbiguousSlotInterval}, since it would
not be deterministic.

A key aspect of the Marlowe semantics is that it checks that a particular
transaction is valid given the current state and contract. Because
transactions are deterministic, there should be no reason why someone
accidentally sends a transaction that is invalid for a given \texttt{State}
and \texttt{Contract}, since it will only result in a cost to that participant.

It is possible that, due to a race condition, a participant will send a
transaction that no longer applies to a running \texttt{Contract} and
\texttt{State}, but such a transaction would simply be ignored by the
blockchain.

@@

This function can be factored into four main functions:
\begin{itemize}
\item \texttt{reduceContractStep} - this function simplifies the topmost
construct that does not require an input to be executed (i.e: anything
but a \texttt{When} that has not expired or a \texttt{Close} when
accounts are empty). It only simplifies the \texttt{When} construct
if it has expired (i.e. the timeout specified in the \texttt{When}
is less than or equal to the minimum slot number). In the case of
the \texttt{Close} contract, it only refunds one of the accounts at
a time. 
\item \texttt{reduceContractUntilQuiescent} - this function calls \texttt{reduceContractStep}
repeatedly until it has no further effect (i.e. it reaches a fixed
point). 
\item \texttt{applyInput} - this function processes one single input.
The topmost construct must be a \texttt{When} that
is expecting that particular input and has not expired. 
\item \texttt{applyAllInputs} - this function processes a list of inputs.
It calls \texttt{applyInput} for each of the inputs in the list, and
calls \texttt{reduceContractUntilQuiescent} before and after every
call to \texttt{applyInput}.
\end{itemize}
The \texttt{State} stores information about the amount of money and
tokens in the contract at a given time, together with the choices
made, \texttt{Let} bindings made, and a lower bound for the current
slot:

@@

Many of the design decisions behind Marlowe have been made with the
aim of preventing potential errors. For example:
\begin{itemize}
\item \textbf{Classification of money and tokens into accounts} separates concerns.
Marlowe will never spend more money or tokens than there are in an
account, even if there is more available in the contract. But when
required to pay more than is in an account, it will pay as much as
is available in the account, in order to remain as close as possible
to the original intention of the contract.
\item \textbf{Account identifiers include an account owner}. An account owner
is a participant that will get the money or tokens remaining in an account
when a contract terminates. At the same time, the only construct that
can pause the execution of a contract is the \texttt{When} construct,
which has a timeout, this ensures that all contracts eventually expire
and terminate. Together, these properties ensure that no money or tokens
are locked in the contract forever.
\item \textbf{No negative deposits or payments}. Marlowe treats negative
amounts in deposits and payments as zero. However, at the same time,
if there happens to be a request for deposit with a negative amount,
it continues gracefully: it will still wait for a null deposit to
be made, and it will continue as if everything is correct; this way,
the expected execution of the contract is disturbed as little as possible.
\item \textbf{Upper limit in the number of inputs} that a Marlowe contract
can accept throughout its lifetime. This limit is implied by the maximum
number of nested \texttt{When} constructs in the contract, since only one input
per \texttt{When} can be accepted. At the same time, transactions that have
no effect on the contract are considered invalid, thus there is also
a limit in the maximum number of transactions a contract can accept
throughout its life. This bound prevents DoS attacks, and it makes static analysis
easier. We discuss this in more detail in Section~\ref{sec:Making-Marlowe-symbolic}.
\end{itemize}

@@

In this section, we briefly present and reflect on a technique that
can be used to convert a concrete implementation of a Haskell function
into a symbolic one by using the SBV library, and to use this symbolic
implementation for static analysis. In particular, we explore this
technique in the context of the Marlowe semantics.

@@

SBV does not currently support complex custom data types. At the time
of writing, SBV has some support for nonary data types, but not for
data types with parameters.
It does provide symbolic versions of the \texttt{Either} type and the \texttt{Tuple} type.
Thus, it is possible to have an isomorphism between custom data types and
nested \texttt{Either} and \texttt{Tuple} type-synonyms. Our original
implementation automatically generates conversion functions between these
by using Template Haskell. By using this technique the symbolic implementation
can remain similar to the concrete one.

@@

In the case of Marlowe, we have natural bounds that we can use because,
given a concrete contract, we can easily infer:
\begin{itemize}
\item The maximum number of inputs that can have an effect on the contract
\item The maximum number of transactions that can have an effect on the
contract
\item All of the account, choice, and \texttt{Let} identifiers that will
be used in the contract
\item The number of participants that will participate in the contract
\end{itemize}
From this data, we can also deduce an upper bound for:
\begin{itemize}
\item The number of times that
\texttt{computeTransaction}, \texttt{reduceContractStep}, and \texttt{applyInput} may
be called.
\item The number of accounts that the contract will use, and an
the number of elements there may be in each of the associative maps
that comprise the \texttt{State} of the contract at any given point
of the execution.
\end{itemize}
If we keep the \texttt{Contract} parameter of the semantics concrete,
we can use it to bound the execution, because every call to \texttt{reduceContractStep}
will either make no progress or remove one of the constructs, with
the exception of \texttt{Close}. But for \texttt{Close} we know we
just have to call it as many times as the maximum number of accounts
there can possibly be, since every time we call it it will refund
one of the accounts.

@@

However, there is one more problem: the output \texttt{Contract} returned
by the function, which is wrapped inside the \texttt{TransactionOutput},
is symbolic, since it depends on the current \texttt{TransactionInput}
and \texttt{State}, which are both symbolic.

@@

\section{Making static analysis efficient\label{sec:Making-static-analysis-efficient}}

In this section, we explain the techniques that we mentioned in Section~\ref{sec:Introduction}
in more detail, and we illustrate those techniques with examples from
the static analysis implementation of Marlowe.

@@

Some of the modifications that we can apply to a static analysis implementation
are localised, and they do not fundamentally affect the overall structure
of the implementation. This means that the implementation can remain
intuitive and close to the original semantics and model that we are
trying to analyze, as well as making changes less likely to introduce
errors than a more serious restructuring.

@@

High level libraries like SBV, and even standards like SMTLib, support
the use and construction of considerably high level abstractions,
e.g: custom data-types, list, sets\dots{} Unfortunately, even though
high level abstractions make reasoning about code easier, they often
prevent certain optimisations, since they abstract out aspects of
the implementation that in our particular case may be concrete.

For example, in the case of Marlowe's static analysis, we initially
implemented a symbolic associative map primitive with the only limitation
that it needed a concrete bound in the number of elements. This is
straightforward to realise using the symbolic implementation of list
and tuple, both provided by SBV. However, this can be very costly:
searching a list symbolically requires constraints that check
the key of every element in the list up to the maximum capacity of
the associative map, because it is not known at the time of symbolic
execution how many elements the list has.

However, in Marlowe we know the values of all the keys that we are
going to use, because the contract is concrete, and only \texttt{Account},
\texttt{Choice}, and \texttt{Let} identifiers that are mentioned in
the contract will ever make it into the \texttt{State}. So we do not
need keys of the associative map to be symbolic, we can use a concrete
associative map with symbolic values.

@@

\subsubsection{Reducing search space by normalising parameters.
\label{subsec:Reducing-search-space-by-normalising-parameters}}

@@

For example, in the case of Marlowe's static analysis,
Marlowe allows several inputs to be combined in one single transaction.
This functionality is important because each transaction requires
the issuer to pay fees, so combining several inputs in a single transaction
can translate into important savings. On the other hand, it also means
that static analysis must consider many more possibilities, since
the number of ways of partitioning inputs in transactions grows exponentially
in the number of inputs.

@@

This optimization reduces the search space considerably, but transactions
may still have no inputs, so it still leaves room for many possible
ways of combining the inputs in transactions. We explain how we reduced
the search space even further in Section~\ref{subsec:Heavyweight-modifications}.

@@

When optimising, if our solution is a local minimum, small changes
to the parameters will not grant any improvement to the value function.
For that reason, in this section, we explore ways of optimizing that
may imply considerable rewriting of our properties, constraints, and
static analysis implementation in general.

\subsubsection{Reducing search space by using normalised execution paths relevant
to the property. \label{subsec:Reducing-search-space-by-using-normalised-execution-paths-relevant-to-the-property}}

One approach we can use that helps improve static analysis efficiency
is to implement the property thinking of the property we want to test
instead of trying to make it as close as possible to the full implementation
of the system. We do not even need to think about the representation
of a counterexample (we will discuss that in the next section), but
only in what are the conditions for the property to be false.

@@

For example, the main property we want to check is whether there is
any possible execution that produces a failed payment. Thus, we look
at executions instead of looking at inputs. The most complicated construct
in terms of executions is the \texttt{When} construct, since it allows
for two transactions that are separate in time to have different effects
depending on when they are issued, all other constructs will get resolved
instantly in one way or another. But, without loss of generality,
we can structure them as Figure~\ref{fig:Distribution-of-transactions}
shows: we can conceptually break the contract tree into subtrees,
where each subtree has a \texttt{When} construct as its root, with
the exception of the subtree that has the same root as the original
contract tree.

@@

Each level of subtrees will correspond to a transaction, i.e: the
root subtree will correspond to the first transaction, the set of
subtrees that are children of the first subtree (in the original tree)
will correspond to the second transaction, and so on. There may be
some paths which do not require as many transactions/subtrees as other
paths, because they go across fewer \texttt{When} constructs before
finding a \texttt{Close} construct.

We can assign one transaction to each path of execution before a \texttt{When},
because if the maximum slot number of the transaction is lower than
the slot number in the timeout and the transaction has no inputs,
then the execution will stop before the \texttt{When}, so we need
to account for that possibility. In other words, if two parts of the
execution path may have different slot numbers when we go through
them, we need to consider them as separate transactions, because they
may have payments that depend on those values.

On the other hand, if the minimum slot number of the transaction is greater
or equal than the slot number in the timeout, then the first transaction
will execute past the \texttt{When} and into the timeout branch, and
we will not need the second transaction. In the same way, if we include
an input that is allowed by the \texttt{When} into the first transaction,
then the first transaction will continue into the corresponding \texttt{When}
branch, and we will not need the second transaction.

@@

One subtle detail that we realised during testing is that, even though
the diagram above shows that a \texttt{When} belongs to the new transaction,
because the \texttt{When} will not be consumed by the previous transaction,
\texttt{Observation}s in the \texttt{Notify} cases will correspond to the
information there was in the \texttt{State} before the \texttt{When}
was executed, with the exception of the slot interval, which will still be
considered to be that of the transaction that triggers the \texttt{When} (the
new transaction). The same is true for \texttt{Value}s in the \texttt{Deposit}
cases, since the amount to deposit must be calculated without considering the
effects of the deposit itself, we need to refer to the \texttt{State} before
the deposit was made.

@@

When we initially implemented the efficient version of static analysis
for Marlowe, we did not pay any attention to the inputs and outputs.
The first version we wrote of the SBV property would not take any
symbolic parameters. Originally, it would simply take a concrete \texttt{Contract}
as input, and it would return a symbolic Boolean that determined whether
the \texttt{Contract} was valid or not. However, if a \texttt{Contract}
turns out to be invalid, we will also want to know why, so we later modified
the property to give a counterexample that illustrated what went wrong.
The original implementation still used some intermediate symbolic
variables, but they were anonymous, and they were created during the
exploration of the contract.

@@

The solution to this increase in execution time was to pass as input
a fixed list of transactions, each of which being represented by a
tuple with four symbolic integers:
\begin{enumerate}
\item An integer representing the minimum slot
\item An integer representing the maximum slot
\item An integer representing the \texttt{When} case whose input is being
included in the transaction, where zero represents the timeout branch
(and there being no inputs to the transaction).
\item An integer representing the amount of money or tokens (if the input
is a \texttt{Deposit}), or the number chosen (if the input is a \texttt{Choice})
\end{enumerate}
If we do not need a transaction, we just set all four numbers to $-1$.

@@

In order to avoid these problems, we combine the use of automated proof
assistants to verify our assumptions and the use of property based testing to
test for consistency and equivalence.

>>>>

\maketitle
\begin{abstract}
SMT solvers can verify properties automatically and efficiently, and they offer increasing
flexibility on the ways those properties can be described. However, the way in which
properties are described can affect the computational cost of verifying them considerably. In addition, it is hard to predict how changes on the way of describing a property
translate into changes to the computational costs.

@@

Thanks to static analysis, we can automatically check beforehand whether any
payments promised by a Marlowe contract can be fulfilled in every
possible execution of the contract. If a Marlowe contract has passed
the static analysis, we will have a very high assurance
that whenever the contract says we will be paid a certain amount of
money, any execution of the contract will indeed have enough money to
make such payment.

@@

This paper contributes a number of approaches that can be used when
optimising static analysis, with examples extracted from a case-study where
we applied these approaches, and an overview of the techniques used
to test the correctness of the optimisations.

The case study presented is that of the optimisation of Marlowe's
static analysis. During this work, the techniques applied allowed
us to reduce the analysis time of an implementation that followed
the semantics closely and took a couple of minutes to analyse contracts
of a few kilobytes, to one where the same contract would take less
than a second, and where a four-person crowdfunding contract that
when fully expanded occupies about 19 megabytes, can be analysed in
around 10 minutes using Z3. In the original implementation, we were
not able to analyse the 19 megabyte contract because the analysis
ran out of memory after half an hour.

@@

modifications are local and can be done without fundamentally changing
the implementation. We consider three main ideas:
\begin{itemize}
\item \textbf{Removing unnecessary parts from the analysis}. If they do not affect
the property that is being verified then we can just remove them from
the implementation and reduce its complexity.
\item \textbf{Avoiding high level abstractions}. High level abstractions aid reasoning
and avoid errors, but also introduce complexity that is not necessarily
needed.
\item \textbf{Reducing the search space by normalising parameters}. If there are several
ways of representing some inputs, and the different representations
have no impact on the analysis, we can remove all but one when
implementing static analysis.
\end{itemize}

@@

modifications are more fundamental approaches that require considerable
changes to the structure of the implementation. We consider two main ideas:
\begin{itemize}
\item \textbf{Reducing the search space by using normalised execution paths relevant
to the property}. Instead of using the search space to model inputs, we
use it to model possible executions, and we ensure that we only
represent each possible group of equivalent executions once, i.e:
executions with the same normal form (or equivalence partitioning).
\item \textbf{Minimizing the representation of inputs and outputs}. We can benefit
from encoding inputs and outputs as concisely as possible, and discarding
information that we can infer in other ways.
\end{itemize}
In the following sections, we introduce the semantics of Marlowe as
a case study (Section~\ref{sec:Marlowe-intro}), we cover a
general approach to static analysis (Section~\ref{sec:Making-Marlowe-symbolic}),
we then explore both lightweight (Section~\ref{subsec:Lightweight-modifications})
and heavyweight (Section~\ref{subsec:Heavyweight-modifications})
optimisation techniques in more detail, and we illustrate them with
examples of how they apply to Marlowe static analysis. In Section~\ref{sec:Testing},
we briefly discuss some ways in which property based testing can be
used to address potential problems that may arise from the implementation
of the heavyweight optimisations described in the previous section.
In Section~\ref{sec:Measurements}, we present empirical results that
show the effects of heavyweight optimisations on the execution time
and memory usage of Marlowe's static analysis.

@@

The Marlowe language is realised as a set of mutually recursive Haskell
data types. Marlowe contracts regulate interactions between a finite number
of participants, determined before the contract is deployed.

Marlowe contracts are able to receive payments, store money and tokens,
ask for input from the participants, and redistribute stored money
and tokens back among participants. The decision about which and when of
these actions are carried out is determined by the contract on the basis
of the information that is available at each point in time. Participants
may correspond to either individual public keys or to tokens (\texttt{Role}s).
In turn, \texttt{Role}s may be controlled by separate independent contracts.

@@

There are 5 constructs of type \texttt{Contract}:
\begin{itemize}
\item \texttt{Close} -- signals the end of the life of a contract. Once
it is reached, all the money and tokens stored in the accounts of
the contract will be refunded to the owner of each of the respective
accounts.
\item \texttt{If} -- immediately decides how the contract must continue
from between two possibilities and depending on a given Boolean condition
that we call an \texttt{Observation}. The value of an \texttt{Observation}
can depend on previous choices made by participants, on amounts of money
and tokens stored in the contract, or other factors.
\item \texttt{Let} -- immediately stores a \texttt{Value} for later use.
Expressions of type \texttt{Value} in Marlowe are evaluated to integer
values and they depend on information available to the contract at the
time of evaluation. For example, a \texttt{Let} construct can be used to
store the amount of money available in a particular account at a given point
in time, for use at a later stage of the contract execution.
\item \texttt{When} -- waits for an external input. The \texttt{When} construct
also specifies a timeout slot\footnote{Slots are blockchain's proxy for time,
they are added to the blockchain periodically.}: after this slot has been reached, the
\texttt{When} construct expires, and no longer accepts any input.
There are three kinds of input:
\begin{itemize}
\item \texttt{Deposit} -- Waits for a participant to deposit an amount
of money or tokens (specified as a \texttt{Value}) in an account of the contract.
\item \texttt{Choice} -- Waits for a participant to make a choice. A choice
is represented as an integer number from a set specified by the contract.
\item \texttt{Notify} -- Waits for an \texttt{Observation} to be true.
Because contracts are reactive (they cannot initiate transactions),
it is necessary for an external actor to \texttt{Notify} the contract
that the \texttt{Observation} has become true. However, this notification
does not have to be carried out by a participant, anybody can do it.
\end{itemize}
\item \texttt{Pay} -- immediately makes a payment between accounts of the
contract, or from an account of the contract to a given participant.
The amount transferred is specified as a \texttt{Value}.
\end{itemize}

@@

As we mentioned in the previous section, Marlowe
contracts are passive: their code is executed as part of the validation of
transactions that are submitted to the blockchain. These transactions
need to be submitted by participants or their representatives (e.g.
user wallets) and their validation is carried out atomically and
deterministically.

Each transaction may include a list of inputs, a set of signatures,
a slot interval (in which the transaction can be executed), a set
of input UTxOs (incoming money
and tokens), and a set of outputs (outgoing money and tokens). For
a transaction to be valid in Marlowe, the transaction must have the
same effect for every slot within that slot range (be deterministic).
For example, if a transaction has a minimum slot number which is before
a timeout, and a maximum slot which after a timeout then the transaction
will fail with the error \texttt{AmbiguousSlotInterval}, since it would
not be deterministic.

A key aspect of the Marlowe semantics is that it checks that a particular
transaction is valid given the current state and contract. Because
transactions are deterministic, there should be no reason why someone
accidentally sends a transaction that is invalid for a given \texttt{State}
and \texttt{Contract}, since it will only result in a cost to that participant.
However, it is still possible that, due to a race condition, a participant will send a
transaction that no longer applies to a running \texttt{Contract} and
\texttt{State}, but such a transaction would simply be ignored by the
blockchain.

@@

This function can be factored into four main functions:
\begin{itemize}
\item \texttt{reduceContractStep} - this function executes the topmost
construct that does not require an input to be executed (i.e: anything
but a \texttt{When} that has not expired or a \texttt{Close} when
accounts are empty). It only simplifies the \texttt{When} construct
if it has expired (i.e. the timeout specified in the \texttt{When}
is less than or equal to the minimum slot number). In the case of
the \texttt{Close} contract, it only refunds one of the accounts at
each invocation. 
\item \texttt{reduceContractUntilQuiescent} - this function calls \texttt{reduceContractStep}
repeatedly until it has no further effect,
\item \texttt{applyInput} - this function processes one single input.
The topmost construct must be a \texttt{When} that
is expecting that particular input and has not expired. 
\item \texttt{applyAllInputs} - this function processes a list of inputs.
It calls \texttt{applyInput} for each of the inputs in the list, and
calls \texttt{reduceContractUntilQuiescent} before and after every
call to \texttt{applyInput}.
\end{itemize}
The \texttt{State} stores information about the amount of money and
tokens in the contract at a given time, together with the choices
made, \texttt{Let} bindings made, and a lower bound for the current
slot:

@@

Many of the design decisions behind Marlowe have been made with the
aim of preventing potential errors. For example:
\begin{itemize}
\item \textbf{Classification of money and tokens into accounts} separates concerns.
A Marlowe will never spend more money or tokens than there are in an
account, even if there is more available in the contract. But when
required to pay more than is in an account, it will pay as much as
is available in the account, in order to remain as close as possible
to the original intention of the contract.
\item \textbf{Account identifiers include an account owner}. An account owner
is a participant that will get the money or tokens remaining in an account
when a contract terminates. At the same time, the only construct that
can pause the execution of a contract is the \texttt{When} construct,
which has a timeout, this ensures that all contracts eventually expire
and terminate. Together, these properties ensure that no money or tokens
are locked in the contract forever.
\item \textbf{No negative deposits or payments}. Marlowe treats negative
amounts in deposits and payments as zero. However, at the same time,
if there happens to be a request for deposit with a negative amount,
it continues gracefully: it will still wait for a null deposit to
be made, and it will continue as if everything is correct; this way,
the expected execution of the contract is disturbed as little as possible.
\item \textbf{Upper limit in the number of inputs} that a Marlowe contract
can accept throughout its lifetime. This limit is implied by the maximum
number of nested \texttt{When} constructs in the contract, since only one input
per \texttt{When} can be accepted. At the same time, transactions that have
no effect on the contract are considered invalid, thus there is also
a limit in the maximum number of transactions a contract can accept
throughout its life. This bound prevents DoS attacks, and it makes static analysis
easier. We discuss this in more detail in Section~\ref{sec:Making-Marlowe-symbolic}.
\end{itemize}

@@

In this section, we briefly present and reflect on a technique that
can be used to convert a concrete implementation of a Haskell function
into a symbolic one by using the SBV library \cite{erkok2019sbv}, and to use this symbolic
implementation for static analysis. In particular, we explore this
technique in the context of the Marlowe semantics.

@@

SBV does not currently support complex custom data types. At the time
of writing, SBV has some support for 0-ary data types, but not for
data types with parameters.
It does provide symbolic versions of the \texttt{Either} type and the \texttt{Tuple} type.
Thus, it is possible to have an isomorphism between custom data types and
nested \texttt{Either} and \texttt{Tuple} type-synonyms. Our original
implementation automatically generates conversion functions between these
by using Template Haskell \cite{sheard2002template}. By using this technique the symbolic
implementation can remain similar to the concrete one.

@@

In the case of Marlowe, we have natural bounds that we can use because,
given a concrete contract, we can easily infer:
\begin{itemize}
\item The maximum number of inputs that can have an effect on the contract
\item The maximum number of transactions that can have an effect on the
contract
\item All of the account, choice, and \texttt{Let} identifiers that will
be used in the contract
\item The number of participants that will participate in the contract
\end{itemize}
From this data, we can also deduce an upper bound for:
\begin{itemize}
\item The number of times that
\texttt{computeTransaction}, \texttt{reduceContractStep}, and \texttt{applyInput} may
be called.
\item The number of accounts that the contract will use, and an
the number of elements there may be in each of the associative maps
that comprise the \texttt{State} of the contract at any given point
of the execution.
\end{itemize}
If we keep the \texttt{Contract} parameter concrete,
we can use it to bound the execution, because every call to \texttt{reduceContractStep}
will either make no progress or remove one of the constructs, with
the exception of \texttt{Close}. But for any particular \texttt{Contract},
we know we only need to call \texttt{reduceContractStep} at most as many times as
the maximum number of accounts there can possibly be, since, every time we call
it, it will refund one of the accounts.

@@

There is one more problem: the output \texttt{Contract} returned
by the function, which is wrapped inside the \texttt{TransactionOutput},
is symbolic, since it depends on the current \texttt{TransactionInput}
and \texttt{State}, which are both symbolic.

@@

\section{Making static analysis more efficient\label{sec:Making-static-analysis-efficient}}

In this section, we explain, in more detail, the techniques that we mentioned
in the introduction, and we illustrate those techniques with
examples from the static analysis implementation of Marlowe.

@@

Some of the modifications that we can apply to a static analysis implementation
are localised, and they do not fundamentally affect the overall structure
of the implementation. This means that the implementation can remain close
to the original semantics and model that we are
trying to analyze, as well as making changes less likely to introduce
errors than a more serious restructuring.

@@

High level libraries like SBV, and even standards like SMTLib, support
the use and construction of high level abstractions,
e.g: custom data-types, list, sets\dots{} Unfortunately, even though
high level abstractions make reasoning about code easier, they often
prevent certain optimisations, since they abstract out aspects of
the implementation that in our particular case may be concrete.

For example, in the case of Marlowe's static analysis, we initially
implemented a \emph{symbolic} associative map primitive with the only limitation
that it needed a concrete bound in the number of elements. This is
straightforward to realise using the symbolic implementation of list
and tuple, both provided by SBV. However, this can be very costly:
searching a list symbolically requires constraints that check
the key of every element in the list up to the maximum capacity of
the associative map, because it is not known at the time of symbolic
execution how many elements the list has.

However, in Marlowe we know the values of all the keys that we are
going to use in maps, because the contract is concrete, and only \texttt{Account},
\texttt{Choice}, and \texttt{Let} identifiers that are mentioned in
the contract will ever make it into the \texttt{State}. So we do not
need keys of the associative map to be symbolic, we can use a concrete
associative map with symbolic values.

@@

\subsubsection{Reducing the search space by normalising parameters.
\label{subsec:Reducing-search-space-by-normalising-parameters}}

@@

For example, in the case of Marlowe's static analysis,
Marlowe allows several inputs to be combined into a single transaction.
This functionality is important because each transaction requires
the issuer to pay fees, so combining several inputs in a single transaction
can translate into important savings. On the other hand, it also means
that static analysis must consider many more possibilities, since
the number of ways of partitioning inputs in transactions grows exponentially
in the number of inputs.

@@

This optimization reduces the search space considerably, but transactions
may still have no inputs, so there are still many ways of distributing
the inputs in transactiongs. We explain how we reduced
the search space even further in Section~\ref{subsec:Heavyweight-modifications}.

@@

When optimising, if our solution is a local minimum, small changes
to the parameters will not grant any improvement to the result.
For that reason, in this section, we explore ways of optimizing that
may imply considerable rewriting of our properties, constraints, and
static analysis implementation in general.

\subsubsection{Reducing the search space by using normalised execution paths relevant
to the property. \label{subsec:Reducing-search-space-by-using-normalised-execution-paths-relevant-to-the-property}}

One approach we can use that helps improve static analysis efficiency
is to focus on the property we want to test instead of trying to make
it as close as possible to the full implementation
of the system. We do not even need to think about the representation
of a counterexample (we will discuss that in the next section), but
only in what are the conditions for the property to be false.

@@

For example, the main property we want to check is whether there is
any possible execution that produces a failed payment. Thus, we look
at executions instead of looking at inputs. The most complicated construct
in terms of executions is the \texttt{When} construct, since it allows
for two transactions that are separate in time to have different effects
depending on when they are issued, all other constructs will get resolved
instantly in one way or another. But, without loss of generality,
we can structure them as Figure~\ref{fig:Distribution-of-transactions}
shows: we can conceptually break the contract tree into subtrees,
where each subtree has a \texttt{When} construct as its root, with
the exception of the subtree at the root of the original tree.

@@

Each level of subtrees will correspond to a transaction, i.e: the
root subtree will correspond to the first transaction, the set of
subtrees that are children of the first subtree (in the original tree)
will correspond to the second transaction, and so on. There may be
some paths which do not require as many transactions/subtrees as other
paths, because they traverse fewer \texttt{When} constructs before
finding a \texttt{Close} construct.

We can assign one transaction to each segment of execution
before starting at a \texttt{When} or the root of the tree,
because if the maximum slot number of the
transaction is lower than the slot number in the timeout and the
transaction has no inputs, then the execution will stop before the
\texttt{When}, so we need to account for that possibility. In other
words, if two parts of the execution path may have different slot
numbers when we go through them, we need to consider them as separate
transactions, because they may have payments that depend on those values.

On the other hand, if the minimum slot number of a transaction is greater
or equal than the slot number in the timeout, then the first transaction
will execute past the \texttt{When} and into the timeout branch, and
we will not need the second transaction. In the same way, if we include
an input that is allowed by the \texttt{When} into the first transaction,
then the first transaction will continue into the corresponding \texttt{When}
branch, and we will not need the second transaction.

@@

One detail that we realised during testing is that, even though
the diagram above shows that a \texttt{When} belongs to the new transaction,
because the \texttt{When} will not be consumed by the previous transaction,
\texttt{Observation}s in the \texttt{Notify} cases will correspond to the
information there was in the \texttt{State} before the \texttt{When}
was executed, with the exception of the slot interval, which will still be
considered to be that of the transaction that triggers the \texttt{When} (the
new transaction). The same is true for \texttt{Value}s in the \texttt{Deposit}
cases, since the amount to deposit must be calculated without considering the
effects of the deposit itself, we need to refer to the \texttt{State} before
the deposit was made.

@@

When we initially implemented the efficient version of static analysis
for Marlowe, we did not pay any attention to the inputs and outputs.
The first version would simply take a concrete \texttt{Contract}
as input, and it would return a symbolic Boolean that determined whether
the \texttt{Contract} was valid or not. However, if a \texttt{Contract}
turns out to be invalid, we will also want to know why, so we later modified
the property to give a counterexample that illustrated what went wrong.
The original implementation still used some intermediate symbolic
variables, but they were anonymous, and they were created during the
exploration of the contract.

@@

The solution to this increase in execution time was to pass as input, to
the static analysis implementation, a fixed list, with each element corresponding
to a transaction and consisting of a tuple with four symbolic integers:
\begin{enumerate}
\item An integer representing the minimum slot
\item An integer representing the maximum slot
\item An integer representing the \texttt{When} case whose input is being
included in the transaction, where zero represents the timeout branch
(and there being no inputs to the transaction).
\item An integer representing the amount of money or tokens (if the input
is a \texttt{Deposit}), or the number chosen (if the input is a \texttt{Choice})
\end{enumerate}
If we do not need a transaction, we just set all four numbers to $-1$.

@@

In order to avoid these problems, we combine the use of automated proof
assistants to verify our assumptions and the use of property based testing to
test for consistency and equivalence (as we describe in
Section~\ref{subsec:Testing-for-consistency}).
